---
title 1980 Census Table
author Rob Kemp
date 11/12/2021
output html_document
---

```{r setup, include=FALSE}
knitropts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(flextable)
library(gt)
```

## Objective

This document provides an example of a table inspired by the 1980 U.S. Decennial Census using data from that census. These tables are unique because of complex headers and subtotals. These data are interesting as a historical look at origin classification.

## Data

Data for this table come from the IPUMS NHGIS project table NT12A and NT13A from the 1980 Decennial Census. These data represent Age totals and those of Spanish Origin (Munson et. al . 2021). 

```{r data clean, echo=FALSE}

nhgis <- read_csv("data/nhgis0009_ds104_1980_state.csv")

age_total <- nhgis%>%
  select(STATE, `Under 5 years`=C7B001, `5-17 years`=C7B002, `18-64 years`=C7B003, `65 years old and over`=C7B004)

age_spanish <- nhgis%>%
  select(STATE, `Under 5 years`=C7D001, `5-17 years`=C7D002, `18-64 years`=C7D003, `65 years old and over`=C7D004)

```

## Age by Spanish Origin by State

```{r table}

# Join data and remove District Of Columbia
data <- age_total%>%
  inner_join(age_spanish, by=c("STATE"="STATE"))%>%
  filter(STATE!="District Of Columbia")%>%
  mutate(seq=2:length(age_total$STATE),
         Total=`Under 5 years.x` + `5-17 years.x` + `18-64 years.x` + `65 years old and over.x`,
         `Under 5 years`=`Under 5 years` - )%>%
  select(STATE, Total, `Under 5 years.x`,`5-17 years.x`,`18-64 years.x`,`65 years old and over.x`,`Under 5 years.y`,`5-17 years.y`,`18-64 years.y`,`65 years old and over.y`)

#Create Totals Row
totals <- data%>%
  summarize(Total=sum(Total), `Under 5 years.x`=sum(`Under 5 years.x`), `5-17 years.x`=sum(`5-17 years.x`), `18-64 years.x`=sum(`18-64 years.x`), `65 years old and over.x`=sum(`65 years old and over.x`), `Under 5 years.y`=sum(`Under 5 years.y`), `5-17 years.y`=sum(`5-17 years.y`), `18-64 years.y`=sum(`18-64 years.y`), `65 years old and over.y`=sum(`65 years old and over.y`))%>%
  mutate(STATE="Total",
         seq=1)%>%
  select(STATE, Total, `Under 5 years.x`,`5-17 years.x`,`18-64 years.x`,`65 years old and over.x`,`Under 5 years.y`,`5-17 years.y`,`18-64 years.y`,`65 years old and over.y`, seq)

# Bind totals to state data and drop seq var
data <- bind_rows(totals, data)%>%
  select(-seq)

# Create column names with spaces after to sub in to make final table show only the age categories.
names=c("State", "Total", "Under 5 years", "5-17 years", "18-64 years", "65 years old and over", "Under 5 years ", "5-17 years ", "18-64 years ", "65 years old and over ")
colnames(data) <- names

table <- data%>%
  flextable()%>%
  add_header_row(colwidths = c(1,1,4,4),
                 values = c("", "The State", "Non-Spanish","Spanish"))%>%
  add_header_row(colwidths = c(2,8), values=c("", "Origin"))%>%
  theme_vanilla()%>%
  vline(j = c('State', '`Total`', '`65 years old and over`', '`65 years old and over `'), border = fp_border_default(), part = "all")

 
# table <- data%>%
#   gt()%>%
#   tab_spanner(
#     label = "Total",
#     columns = c(`Under 5 years`, `5-17 years`, `18-64 years`, `65 years old and over`)
#   )%>%
#   tab_spanner(
#     label = "Spanish Origin",
#     columns = c(`Under 5 years `, `5-17 years `, `18-64 years `, `65 years old and over `)
#   )%>%
#   summary_rows(columns=`Under 5 years`:`65 years old and over `, fns=list(Total="sum"))
# table

```

## References

Steven Manson, Jonathan Schroeder, David Van Riper, Tracy Kugler, and Steven Ruggles. IPUMS National Historical Geographic Information System: Version 16.0 [dataset]. Minneapolis, MN: IPUMS. 2021. http://doi.org/10.18128/D050.V16.0